---
title: Dynamics of cellulose degradation by soil microorganisms from two contrasting
  soil types
author: "Grigory Gladkov"
date: "2024-07-15"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 5
    number_sections: true
    theme: lumen
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
rmarkdown::find_pandoc(dir = "/home/gladkov/.conda/envs/bi/bin")
knitr::opts_knit$set(root.dir = "~/storage/cellulolit/paper/")
options(getClass.msg=FALSE)
source('functions.R', local = knitr::knit_global())
list.files()

```


```{r}

library(tidyverse)
library(phyloseq)
library(igraph)
library(SpiecEasi)

```


```{r, include=FALSE}

setwd("~/storage/cellulolit/paper/")
list.files()

```


```{r}

psits.f <- readRDS("psits.f")
ps.f <- readRDS("ps.f")

```

## 16S basic EDA

```{r}

plot_rich_reads_samlenames_lm(ps.f, group = "Repeat", label = "Repeat")

```

```{r, fig.height=4.5, fig.width=10}

p1 <- plot_alpha_w_toc_2(ps.f, group="Repeat", metric="Observed")
p2 <- plot_alpha_w_toc_2(ps.f, group="Repeat", metric="Shannon")
p3 <- plot_alpha_w_toc_2(ps.f, group="Repeat", metric="InvSimpson")

p_alpha <- ggpubr::ggarrange(p1, p2, p3, nrow = 1)
p_alpha

```

```{r}

amp <- phyloseq_to_ampvis2(ps.f)
amp

```


```{r}

ampvis2::amp_heatmap(amp,
            tax_show = 22,
            group_by = "Day",
            facet_by = "Substrate",
            tax_aggregate = "Phylum",
            tax_class = "Proteobacteria",
            normalise=TRUE,
            showRemainingTaxa = TRUE)

```


```{r, fig.height=12, fig.width=8}

ampvis2::amp_heatmap(amp,
            tax_show = 60,
            group_by = "Day",
            facet_by = "Substrate",
            tax_aggregate = "Genus",
            tax_add = "Phylum",
            normalise=TRUE,
            plot_values_size = 3, 
            showRemainingTaxa = TRUE)

```


```{r, fig.height=12, fig.width=8}

ampvis2::amp_heatmap(amp %>% ampvis2::amp_filter_taxa("Bacteroidota", normalise = TRUE),
            tax_show = 40,
            group_by = "Day",
            facet_by = "Substrate",
            tax_aggregate = "Species",
            tax_add = "Genus",
            normalise=FALSE,
            plot_values_size = 3, 
            showRemainingTaxa = TRUE)

```

## its2S basic EDA

Previosly some sample was removed, based on this picture('its2_C14.4') \
Isn't shown here, because crean code should stay clean.

```{r}

plot_rich_reads_samlenames_lm(psits.f, 'Repeat')

```


## final beta`s plot

```{r, fig.height= 5.7, fig.width=10.2}

p.beta.16 <- beta_custom_norm_NMDS_elli_w(ps.f, Group = "Repeat", Color = "Substrate")
p.beta.its <- beta_custom_norm_NMDS_elli_w(psits.f, Group = "Repeat", Color = "Substrate")

p.beta.16.mod <- p.beta.16 + 
  ggtitle('16S', subtitle = 'NMDS - Bray-Curtis dissimilarity') +
      scale_colour_viridis_d(option = "magma", 
                         aesthetics = "color", 
                         begin = 0.8, 
                         end = 0.001)
p.beta.its.mod <- p.beta.its +
  ggtitle('ITS2', subtitle = 'NMDS - Bray-Curtis dissimilarity') +
      scale_colour_viridis_d(option = "magma", 
                         aesthetics = "color", 
                         begin = 0.8, 
                         end = 0.001)
  
fig.2 <- ggarrange(p.beta.16.mod, p.beta.its.mod)
fig.2
# ggsave(fig.2, filename="../plots_supp/fig2.pdf", width = 10.2, height = 5.7)

```


```{r}

ampits <- phyloseq_to_ampvis2(psits.f)
ampits

```

```{r, fig.height=6, fig.width=7}

ampvis2::amp_heatmap(ampits,
            tax_show = 40,
            group_by = "Day",
            facet_by = "Substrate",
            tax_aggregate = "Phylum",
            # tax_add = "Phylum",
            normalise=TRUE,
            showRemainingTaxa = TRUE)

```

```{r, fig.height=4.5, fig.width=10}


p1 <- plot_alpha_w_toc_2(psits.f, group="Repeat", metric="Observed")
p2 <- plot_alpha_w_toc_2(psits.f, group="Repeat", metric="Shannon")
p3 <- plot_alpha_w_toc_2(psits.f, group="Repeat", metric="InvSimpson")

p_alpha <- ggpubr::ggarrange(p1, p2, p3, nrow = 1)
p_alpha


```

Only Ascomycota

```{r, fig.height=7, fig.width=10}

ampits.nematode <- ampvis2::amp_filter_taxa(ampits, "Ascomycota" ,normalise = TRUE)
ampvis2::amp_heatmap(ampits.nematode,
            tax_show = 40,
            group_by = "Day",
            facet_by = "Substrate",
            tax_aggregate = "Genus",
            # tax_add = "Genus",
            normalise=FALSE,
            showRemainingTaxa = TRUE)

```

Only Nematoda

```{r, fig.height=6, fig.width=10}

ampits.nematode <- ampvis2::amp_filter_taxa(ampits, "Nematoda" ,normalise = TRUE)
ampvis2::amp_heatmap(ampits.nematode,
            tax_show = 40,
            group_by = "Day",
            facet_by = "Substrate",
            tax_aggregate = "Species",
            tax_add = "Genus",
            normalise=FALSE,
            showRemainingTaxa = TRUE)

```

## data normalisation and filtration

```{r, include=FALSE}

out.f2.a <- readRDS('out.f2.a')
out.f2.c <- readRDS('out.f2.c')

netfull.a <- readRDS('netfull.a')
netfull.c <- readRDS('netfull.c')

```

```{r}

ps.ff <- phyloseq::filter_taxa(ps.f, function(x) sum(x > 5) > (0.1*length(x)), TRUE)

ps.f.a <- prune_samples(sample_data(ps.f)$Substrate %in% c('A'), ps.f) 
ps.f.a  <- prune_taxa(taxa_sums(ps.f.a) > 0, ps.f.a)   
ps.ff.a <- phyloseq::filter_taxa(ps.f.a, function(x) sum(x > 5) > (0.1*length(x)), TRUE)

ps.f.c <- prune_samples(sample_data(ps.f)$Substrate %in% c('C'), ps.f) 
ps.f.c  <- prune_taxa(taxa_sums(ps.f.c) > 0, ps.f.c)   
ps.ff.c <- phyloseq::filter_taxa(ps.f.c, function(x) sum(x > 5) > (0.1*length(x)), TRUE)

ps.f2.a <- phyloseq::filter_taxa(ps.f.a, function(x) sum(x > 5) > (0.1*length(x)), TRUE)
ps.f2.c <- phyloseq::filter_taxa(ps.f.c, function(x) sum(x > 5) > (0.1*length(x)), TRUE)

```



```{r, eval=FALSE, cache=TRUE}

out.f2.a <-  ANCOMBC::ancombc2(data=ps.f2.a, 
  fix_formula = "Repeat",
  prv_cut = 0,
  p_adj_method = "BH",
  tax_level = NULL,
  rand_formula= NULL,
  pseudo = 0, 
  pseudo_sens = FALSE,
  s0_perc = 0.05,
  group = "Repeat",
  struc_zero = FALSE,
  neg_lb = TRUE,
  alpha = 0.05,
  n_cl = 20, 
  verbose = TRUE,
  global = FALSE, 
  pairwise = FALSE, 
  dunnet = FALSE, 
  trend = FALSE
  )

out.f2.c <-  ANCOMBC::ancombc2(data=ps.f2.c, 
  fix_formula = "Repeat",
  prv_cut = 0,
  p_adj_method = "BH",
  tax_level = NULL,
  rand_formula= NULL,
  pseudo = 0, 
  pseudo_sens = FALSE,
  s0_perc = 0.05,
  group = "Repeat",
  struc_zero = FALSE,
  neg_lb = TRUE,
  alpha = 0.05,
  n_cl = 20, 
  verbose = TRUE,
  global = FALSE, 
  pairwise = FALSE, 
  dunnet = FALSE, 
  trend = FALSE
  )


```


```{r}

ps.f2.a.anc <- norm_anc(ps.f2.a, out.f2.a)
ps.f2.c.anc <- norm_anc(ps.f2.c, out.f2.c)

beta_custom_norm_NMDS_elli_w(ps.f2.a.anc, Color = "Repeat", Group = "Repeat")
beta_custom_norm_NMDS_elli_w(ps.f2.c.anc, Color = "Repeat", Group = "Repeat")

```



```{r}

ps.f2.a.anc <- norm_anc(ps.f2.a, out.f2.a)
ps.f2.c.anc <- norm_anc(ps.f2.c, out.f2.c)


```

## 16S network analysis

```{r, eval=FALSE}

netfull.a <- SpiecEasi::spiec.easi(ps.f2.a, method='mb', pulsar.params=list(rep.num=500))
netfull.c <- SpiecEasi::spiec.easi(ps.f2.c, method='mb', pulsar.params=list(rep.num=500))

```


```{r}

igfull.a <- SpiecEasi::adj2igraph(getRefit(netfull.a), vertex.attr=list(name=taxa_names(ps.f2.a)))
igfull.c <- SpiecEasi::adj2igraph(getRefit(netfull.c), vertex.attr=list(name=taxa_names(ps.f2.c)))

modfull.a <- cluster_fast_greedy(igfull.a)
modfull.c <- cluster_fast_greedy(igfull.c)

ig.a.mod <- igfull.a
V(ig.a.mod)$color <-  modfull.a$membership

ig.c.mod <- igfull.c
V(ig.c.mod)$color <-  modfull.c$membership

print(modfull.a)
print(modfull.c)

```



```{r}

wtc.ig.a.mod <- cluster_walktrap(ig.a.mod)
wtc.ig.c.mod <- cluster_walktrap(ig.c.mod)

modularity(wtc.ig.a.mod)
modularity(wtc.ig.c.mod)

```

7th group from the chernozem

```{r,  fig.height= 6, fig.width=7}

groups.c <- data.frame('ID' = modfull.c$names, 'Group' = modfull.c$membership)
groups.a <- data.frame('ID' = modfull.a$names, 'Group' = modfull.a$membership)

id.7 <- ps.f2.c@tax_table %>% 
  data.frame() %>% 
  rownames_to_column('ID') %>% 
  right_join(groups.c) %>% 
  filter(Group %in% c('7')) %>% 
pull('ID')

amp.ff.c <- phyloseq_to_ampvis2(ps.f2.c)
amp.ff.c.7 <- ampvis2::amp_filter_taxa(amp.ff.c, tax_vector = id.7, normalise = TRUE)


ampvis2::amp_heatmap(amp.ff.c.7,
            tax_show = 30,
            tax_aggregate = "OTU",
            tax_add = "Genus",
            group_by = "Day",
            plot_values_size = 4,
            showRemainingTaxa = TRUE, normalise = FALSE)

```

### creating net`s plots and treeplots

```{r, fig.height=4, fig.width=8}

library(GGally)

set.seed(1516)

pl_pal <-  viridis::magma(n = 9) 
# option = "magma", aesthetics = "color", begin = 0.8, end = 0.001

p.net.a <- ggnet2(ig.a.mod, 
        size ="0.7",
       node.color = "color", 
       label = FALSE, 
       node.size = 2, 
       label.size = 1,
       mode = "kamadakawai") + 
  guides(color=guide_legend(title="color"), size = FALSE)  + 
  scale_color_manual(values = pl_pal) +
    theme(legend.position="none")  + 
  ggtitle('A', subtitle = 'modularity - 0.28')

p.net.c <- ggnet2(ig.c.mod, 
      size ="0.7",
       node.color = "color", 
       label = FALSE, 
      node.size = 2, 
       label.size = 1,
       mode = "kamadakawai") + 
  guides(color=guide_legend(title="Groups"), size = FALSE)  + 
  scale_color_manual(values = pl_pal)  +
    theme(legend.position="none")  + 
  ggtitle('C', subtitle = 'modularity - 0.33')

fig.3 <- ggarrange(p.net.a, p.net.c)
fig.3
ggsave(fig.3, filename="../plots_supp/fig3.pdf", width = 8, height = 4)

```

Groups taxonomy for A

```{r, fig.height=8, fig.width=10}

tx <- left_join(groups.a, ps.f2.a.anc@tax_table@.Data %>% data.frame() %>% rownames_to_column('ID')) %>% 
  add_column(abnd = taxa_sums(ps.f2.a.anc))


fig.4 <- tx %>% 
  mutate(Phylum = case_when(Phylum %in% "Proteobacteria" ~ Class,
                            !Phylum %in% "Proteobacteria" ~ Phylum) %>% as.factor()) %>% 
  select(c("Genus", "Phylum", "Group", "abnd")) %>% 
  mutate(Group = as.factor(Group)) %>% 
  group_by(Genus, Phylum, Group) %>%
  mutate(Genus = str_replace_na(Genus, replacement = ' ') %>% as.factor()) %>% 
  summarise(area = sum(abnd)) %>% 
  mutate_if(is.character, as.factor) %>% 
  drop_na() %>% 
  filter(Group %in% c('1', '2', '4', '8')) %>% 
  ggplot(aes(area = area, 
             subgroup = Phylum, 
             label = Genus,
             fill = Group)) +
  treemapify::geom_treemap() +
  treemapify::geom_treemap_subgroup_border(color = "white") +
  treemapify::geom_treemap_subgroup_text(place = "centre", 
                                         grow = T, 
                                         alpha = 0.7, 
                                         colour = "black", 
                                         fontface = "italic", 
                                         min.size = 2) +
  treemapify::geom_treemap_text(colour = "white", 
                                place = "topleft", 
                                grow = T, 
                                reflow = T, 
                                layout = 'squarified',
                                min.size = 4) +
  facet_wrap(~Group, labeller = labeller(Group = 
    c("1" = "Group1, rel=9.16%, ASVs=84",
      "2" = "Group2, rel=30.97%, ASVs=164",
      "4" = "Group4, rel=29.17%, ASVs=141",
      "8" = "Group8, rel=25.06%, ASVs=160"))) +
  scale_colour_viridis_d(option = "plasma", 
                         aesthetics = "fill", 
                         begin = 0.2, 
                         end = 0.7) +
  theme(legend.position="bottom") 

fig.4

```

Groups taxonomy for C

```{r, fig.height=8, fig.width=10}

tx <- left_join(groups.c, ps.f2.c.anc@tax_table@.Data %>% data.frame() %>% rownames_to_column('ID')) %>% 
  add_column(abnd = taxa_sums(ps.f2.c.anc))

fig.5 <- tx %>% 
  mutate(Phylum = case_when(Phylum %in% "Proteobacteria" ~ Class,
                            !Phylum %in% "Proteobacteria" ~ Phylum) %>% as.factor()) %>% 
  select(c("Genus", "Phylum", "Group", "abnd")) %>% 
  mutate(Group = as.factor(Group)) %>% 
  group_by(Genus, Phylum, Group) %>%
  mutate(Genus = str_replace_na(Genus, replacement = ' ') %>% as.factor()) %>% 
  summarise(area = sum(abnd)) %>% 
  mutate_if(is.character, as.factor) %>% 
  drop_na() %>% 
  filter(Group %in% c('1', '2', '4', '6','7','9')) %>% 
  ggplot(aes(area = area, 
             subgroup = Phylum, 
             label = Genus,
             fill = Group)) +
  treemapify::geom_treemap() +
  treemapify::geom_treemap_subgroup_border(color = "white") +
  treemapify::geom_treemap_subgroup_text(place = "centre", 
                                         grow = T, 
                                         alpha = 0.7, 
                                         colour = "black", 
                                         fontface = "italic", 
                                         min.size = 2) +
  treemapify::geom_treemap_text(colour = "white", 
                                place = "topleft", 
                                grow = T, 
                                reflow = T, 
                                layout = 'squarified',
                                min.size = 4) +
  facet_wrap(~Group, labeller = labeller(Group = 
    c("1" = "Group1, rel=8.44%, ASVs=40",
      "2" = "Group2, rel=21.35%, ASVs=58",
      "4" = "Group4, rel=15.66%, ASVs=76",
      "6" = "Group6, rel=10.94%, ASVs=26",
      "7" = "Group7, rel=18.25%, ASVs=55",
      "9" = "Group9, rel=19.66%, ASVs=68" ))) +
  scale_colour_viridis_d(option = "plasma", 
                         aesthetics = "fill", 
                         begin = 0.2, 
                         end = 0.7) +
  theme(legend.position="bottom") 

fig.5

ggsave(fig.4, filename="../plots_supp/fig4.pdf", width = 10, height = 8)
ggsave(fig.5, filename="../plots_supp/fig5.pdf", width = 10, height = 8)

```

```{r,  fig.height= 13, fig.width=10}

groups.a <- data.frame('ID' = modfull.a$names, 'Group' = modfull.a$membership)

id.5 <- ps.f2.a@tax_table %>% 
  data.frame() %>% 
  rownames_to_column('ID') %>% 
  right_join(groups.a) %>% 
  filter(Group %in% c('4')) %>% 
  pull('ID')

amp.ff.a <- phyloseq_to_ampvis2(ps.f2.a.anc)
amp.ff.a.5 <- ampvis2::amp_filter_taxa(amp.ff.c, tax_vector = id.5, normalise = TRUE)

ampvis2::amp_heatmap(amp.ff.a.5,
            tax_show = 100,
            tax_aggregate = "OTU",
            tax_add = "Genus",
            group_by = "Time",
            plot_values_size = 4,
            showRemainingTaxa = TRUE, normalise = FALSE)

```


### adding aspiration plots

```{r, fig.height=5.5, fig.width=9}

ps.f.r <- rarefy_even_depth(ps.f, rngseed = 1433) 
er <- estimate_richness(ps.f.r)

p.alpha.16 <- estimate_richness(ps.f.r) %>% 
  rownames_to_column("ID") %>% 
  mutate(ID = str_sub(ID, 2)) %>% 
  left_join(ps.f.r@sam_data %>%  
              data.frame() %>%  
              rownames_to_column("ID") %>% 
              mutate(Repeat = as.factor(Repeat))) %>% 
  group_by(Substrate) %>% 
  pivot_longer(c( "Observed", "Shannon", "InvSimpson", "Respiration")) %>% 
  mutate(Index = factor(name, levels = c("Observed", "Shannon", "InvSimpson", "Respiration"))) %>% 
  ggplot(aes(y = value, x = Day, group = Substrate)) +
    # geom_line(aes(color = name),
    #         alpha = 1) +
    stat_summary(aes(y = value, color = Substrate), fun.y=mean, geom="line") +
    labs(x = "day",
       y = "index value") +  
  labs(title = "16S") + 
  geom_point(aes(shape = Index, color = Substrate)) +
  facet_wrap(~ Index, nrow = 2, scales = "free") +
    scale_colour_viridis_d(option = "magma", 
                         aesthetics = "color", 
                         begin = 0.8, 
                         end = 0.001) +
  theme_bw()

psits.f.r <- rarefy_even_depth(psits.f, rngseed = 1433) 
er.its <- estimate_richness(ps.f.r)

p.alpha.its <- estimate_richness(psits.f.r) %>% 
  rownames_to_column("ID") %>% 
  mutate(ID = str_sub(ID, 1)) %>% 
  left_join(psits.f.r@sam_data 
            %>%  data.frame() 
            %>%  rownames_to_column("ID")
            %>% mutate(Repeat = as.factor(Repeat))) %>% 
  left_join(ps.f.r@sam_data 
            %>%  data.frame() 
            %>%  rownames_to_column("ID")
            %>% dplyr::select(c("Repeat", "Respiration"))
            %>% mutate(Repeat = as.factor(Repeat))) %>% 
  group_by(Substrate) %>% 
  pivot_longer(c( "Observed", "Shannon", "InvSimpson", "Respiration")) %>% 
  mutate(Index = factor(name, levels = c("Observed", "Shannon", "InvSimpson", "Respiration"))) %>% 
  mutate(Substrate = as.character(Substrate)) %>%
  ggplot(aes(y = value, x = Day, group = Substrate)) +
    stat_summary(aes(y = value, color = Substrate), fun.y=mean, geom="line") +
    labs(x = "day",
       y = "index value") +
  labs(title = "ITS2") + 
  geom_point(aes(color = Substrate, shape = Index)) +
  facet_wrap(~ Index, nrow = 2, scales = "free") +
    scale_colour_viridis_d(option = "magma", 
                         aesthetics = "color", 
                         begin = 0.8, 
                         end = 0.001) +
  theme_bw() +
  theme(legend.position = "bottom")

g_legend<-function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)}

mylegend <- g_legend(p.alpha.its)


fig.1 <- gridExtra::grid.arrange(gridExtra::arrangeGrob(p.alpha.16 + theme(legend.position="none"),
                         p.alpha.its + theme(legend.position="none"),
                         nrow=1),
             mylegend, nrow=2,heights=c(12, 1))


fig.1

ggsave(fig.1, filename="../plots_supp/fig1.pdf", width = 9, height = 5.5)

```

### formal stats

```{r}

dist <- phyloseq::distance(ps.f, "euclidean")
metadata <- as(sample_data(ps.f@sam_data), "data.frame")
vegan::adonis2(dist ~ Substrate + Time, data = metadata)

```

```{r}

dist <- phyloseq::distance(ps.f, "euclidean")
metadata <- as(sample_data(ps.f@sam_data), "data.frame")
vegan::adonis2(dist ~ Substrate + Time, data = metadata)

```

```{r}

dist <- phyloseq::distance(psits.f, "euclidean")
metadata <- as(sample_data(psits.f@sam_data), "data.frame")
vegan::adonis2(dist ~ Substrate + Time, data = metadata)

```

```{r}

sessionInfo()

```


```{r}



```



